% This thesis, written during the fall semester 2023, is the finalization of the course TMA4500 Specialization Project. The project constitutes 15 ECTS and focuses on introducing a Bayesian variable importance measure. During the spring semester 2024, I will further extend the work presented in this thesis through a 30 ECTS master’s thesis.
% In the process of creating this thesis, some tools that use artificial intelligence have been used. Mainly, I have used OpenAI’s large language model ChatGPT to help me solve problems I encountered when programming, get input on the structure and flow of my text and some help with grammar and spelling. Most importantly, I wish to stress that all text and findings in this thesis are my own work, and that I have not used artificial intelligence tools to generate any content.
% I want to thank my supervisor Stefanie Muff for her excellent guidance and sup- port throughout the project. I look forward to continuing our work during the spring of 2024.
This masters thesis concludes the Master of Science degree obtained from the program Physics and Mathematics, with a specialization in Industrial Mathematics, at the Norwegian University of Science and Technology (NTNU). In combination with the project thesis \citep{Arnstad:Relative_variable_importance_in_Bayesian_linear_mixed_models:2024}, the masters thesis constitutes 45ECTS, and has been developed during the spring of 2024.
\\
\\
First and foremost I want to thank my supervisor Stefanie Muff, who has been critical in developing the thesis and has provided excellent guidance. For their generous help and creative discussions, I want to thank Javier Aguilar at the University of Stuttgart. I also want to express my gratitude to my fellow students at the study program Physics and Mathematics, who I have become close friends with and have learned a lot from. My time at NTNU has been fantastic, and something I will cherish for the rest of my life. Lastly, I want to thank my family and Emma, who have supported me throughout my studies. You have all been great people to have around, and I am forever grateful. 
\\
\\
Kom igjen Troilljan!